Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models

### 2017-08-04 revised-chapter.tex
Changes to figures

The Earth's magnetosphere is a region around the planet where the Earth's own magnetic field dominates the magnetic field of outer space. It is a region which is impinged upon constantly by the solar wind. The ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere. It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.

Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions. Their values are often calculated by averaging or combining a number of readings taken by instruments, usually magnetometers, around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomenon. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citep{Bartels}. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}. A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurrence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure. 
\end{enumerate}

Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forecasts for various geomagnetic time series. The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).

@@

\subsection{Probabilistic Forecasting}
%Talk about need for probabilistic forecasts.
Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$. One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.

@@

In order to uniquely define the distribution of $\mathbf{f}$, it is required to specify $\mathbf{m}$ and $\mathbf{K}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.

@@

The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The process of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.

@@

\subsection{Model Selection: Hyper-parameters}

Given a GPR model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values for the kernel hyper-parameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$. In order to assign a value to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model. Since GP models are constructed from assumptions about the conditional probability distribution of the data $p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.

@@

The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above. In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$.

The model selection problem has historically received less attention due to a combination of difficulty and numerous design decisions like choice of hyper-parameter optimization techniques, which makes model selection as much of an art as science. 
Choosing a suitable heuristic or algorithm for model selection for GP models is quite often a balancing exercise between the computational limitations of the modeler and the need to explore the hyper-parameter space in a manner that yields a satisfactory predictive model. Some of the techniques used for model selection in the context of GPR include.

@@

Apart from having a practical method for choosing values of kernel hyper-parameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models. One might argue that we group these quantities with the kernel hyper-parameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.

The value of kernel hyper-parameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively. Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyper-parameters which yield a good fit.

Not only do the model orders influence the value of kernel hyper-parameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.

@@

We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}. In order to find appropriate values of the hyper-parameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods. We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection. Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.

@@

\item \textbf{Model Order Selection}: In order to find appropriate values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$, solar wind speed and IMF $B_z$ respectively, create models upto a certain maximum time lag i.e. $3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX. For each such model perform the model selection routine to select values of the kernel hyper-parameters and evaluate the resultant model on the validation set. Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.

\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyper-parameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance. In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation. The GP based OSA prediction models are compared against the naive \emph{Persistence} model which serves as a performance baseline for the OSA prediction of $Dst$.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 19:00 to 2003/06/19 03:00}
\label{fig:ComparePred1}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/11/13 00:00 to 1998/11/15 04:00}
\label{fig:ComparePred2}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm9.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1999/01/13 16:00 to 1999/01/14 20:00}
\label{fig:ComparePred3}
\end{figure}

>>>>

The Earth's magnetosphere is the region above the ionosphere that is defined by the extent of the Earth's magnetic field in space. It extends several tens of thousands of kilometers into space, a region which is impinged upon constantly by the solar wind. Ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere. It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.

Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions. Their values are often calculated by averaging or combining a number of readings taken by instruments, usually magnetometers, around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomenon. Some common indices are the $Kp$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $Kp$: The Kp-index is a discrete valued global geomagnetic
      activity index and is based on 3 hour measurements of the
      K-indices \citep{Bartels}. The K-index itself is a three hour
      long quasi-logarithmic local index of the geomagnetic activity,
      relative to a quiet day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to
      provide a global, quantitative measure of auroral zone magnetic
      activity produced by enhanced ionospheric currents flowing below
      and within the auroral oval \citep{AEIndex}. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the
      magnitude of the magnetic field induced by near equatorial ring
      currents \citep{DesslerAndParker}. A large negative value of
      $Dst$ ($ \leq -100 \ nT$) indicates occurrence of geomagnetic
      storm events which are of particular importance in space weather
      prediction due to their adverse effects on telecommunications
      infrastructure, navigation systems, power grids, satellites etc. 
\end{enumerate}

Space weather forecasting systems usually use in-situ measurements of
solar wind parameters, taken by satellites, as well as historical data
of indices to produce forecasts for various geomagnetic time
series. Real time data of solar wind parameters are recorded by the \emph{ACE}
and \emph{DISCOVR} missions. Geomagnetic indices are often recorded in
ground based measurements. The the \emph{Space Physics Data Facility}
(SPDF) hosted by NASA is a gateway for accessing data from a variety
of sources, including those mentioned above.

@@

\subsection{Probabilistic Forecasting}
%Talk about need for probabilistic forecasts.ny
Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$. One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.

@@

In order to uniquely define the distribution of $\mathbf{f}$, it is
required to specify the mean $\mathbf{m}$ and the covariance matrix
$\mathbf{K}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during substorms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.

@@

The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}. Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The process of assigning values to the \emph{hyperparameters} is crucial in the model building process and is known as \emph{model selection}.

@@

\subsection{Model Selection: Hyperparameters}

Given a GPR model with a kernel function $K_\theta$, the problem of
model selection consists of finding appropriate values for the kernel
hyperparameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.  In order to assign a value to $\theta$, we must
define an objective function ($Q(\theta)$) which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model. Since GP models are constructed from assumptions about the conditional probability distribution of the data $p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.

@@

The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above. In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyperparameters to inform our choice for $\theta$.

The model selection problem has historically received less attention
due to a combination of difficulty and numerous design decisions like
choice of hyper-parameter optimization techniques, which makes model
selection as much of an art as science.

Choosing a suitable heuristic or algorithm for model selection for GP
models is quite often a balancing exercise between the computational
limitations of the modeler and the need to explore the hyper-parameter
space in a manner that yields a satisfactory predictive model. Some of
the techniques used for model selection in the context of GPR include
the following.

@@

Apart from having a practical method for choosing values of kernel hyperparameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models. One might argue that we group these quantities with the kernel hyperparameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.

The value of kernel hyperparameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively. Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyperparameters which yield a good fit.

Not only do the model orders influence the value of kernel hyperparameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.

@@

We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}. In order to find appropriate values of the hyperparameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods. We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection. Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.

@@

\item \textbf{Model Order Selection}: In order to find appropriate values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$, solar wind speed and IMF $B_z$ respectively, create models upto a certain maximum time lag i.e. $3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX. For each such model perform the model selection routine to select values of the kernel hyperparameters and evaluate the resultant model on the validation set. Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.

\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyperparameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance. In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation. The GP based OSA prediction models are compared against the naive \emph{Persistence} model which serves as a performance baseline for the OSA prediction of $Dst$.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 to 2003/06/19}
\label{fig:ComparePred1}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm16.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2012/03/08 to 2012/03/10}
\label{fig:ComparePred2}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm46.png}
\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/11/20 to 2003/11/22}
\label{fig:ComparePred3}
\end{figure}
